Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-08-20 paper.tex
Update on Overleaf.

Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}. Therefore, observing a system at different points in time does not equate to observing different system at a single point in time. This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation.

@@

@@

\paragraph{Platform constraints and docking challenges} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space. For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements. Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space. This issue is part of the docking challenge \citep{Axtelletal1996} (i.e. checking if two models can produce the same results), but more generally, it involves a description of the model and its spatial requirements more detailed than what is currently the rule.

% \comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]

@@

\paragraph{Reproducibility and Applicability} 
%\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}

@@

%\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
%\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}

@@

@@

>>>>

Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}. Therefore, observing a system at different points in time does not equate to observing \added{a} different system at a single point in time. This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation.

@@

\todo{add code and data availability statement here}

@@

\added{\cite{kuhnert2005comparing} map comparison techniques}

\added{\cite{kruskal1983overview} parallel with time-series comparison techniques (see reviewer comment - however more constrained as two dimensional indexing)}

\paragraph{Platform constraints and docking challenges}

An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space. For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements. Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space. This issue is part of the docking challenge \citep{Axtelletal1996} (i.e. checking if two models can produce the same results), but more generally, it involves a description of the model and its spatial requirements more detailed than what is currently the rule.

@@

\paragraph{Reproducibility and Applicability}

@@

@@

\subsection*{\added{Additional figures for the Schelling model}}

\added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text. In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.}

@@

\subsection*{\added{Pseudo-code for models}}

\added{We give below the pseudo-code for the implementations we used of both models.}
